\chapter{Reading Passages}
\label{appendix:reading_passages}

\section{Friends}

Participants first read this short, neutral, three-sentence passage to evaluate their reading level and comfort. The passage was used by Labov in New York and did not target any specific features \citep[417]{labov_2006}.

\begin{quotation}
    When I was nine or ten, I had a lot of friends who used to come over to my house to play. I remember a kid named Henry who had very big feet, and I remember a boy named Billy who had no neck, or at least none to look at. He was a funny kid, all right.
\end{quotation}

\section{The Cat and the Mice}

This adaptation of one of Aesop's Fables was written by Alicia Wassink to specifically target features known to be variable in Washington. The text was taken from \citet{freeman_2014}.

\begin{quotation}
    Once upon a time, a cat passed by a big house that was full of mice. Ever since she was a kitten, the cat had thought to herself, “I would be happy in that home.” So, the very next evening, she moved in with the family that lived in the house. 
    
    Then, the fight between the cat and the mice began. At first, the mice avoided the cat like the plague, huddling under the deck and trying not to make a peep. But the cat simply hid in a corner behind a sack of bagels and waited calmly for them to take the bait. The mice zigged and zagged down the hall, but they only had a vague idea where the cat was hidden. 
    
    When she caught one, she would pen the mouse up in the corner between her paws. The mice would beg for mercy and kick their legs with all their strength. Their nagging cries made quite a din. But, the cat would simply pin them against the wall of her den and eat them up anyway. 
    
    This went on for six days, until finally the mice couldn’t stand it anymore. Feeling angry, they decided to go into their holes and stay there for a week. This way, bragged the mice, the mean old cat would never catch them. “That’s not fair,” said the cat to herself, as she drank the pool of milk in her bowl. “Now the only thing to do is to pull them out by a trick — then I can bake a fine mouse pie and eat like a king again.” So, she thought for a while, paced the length of the room, and hatched a plan. 
    
    Eager to test her new scheme, she climbed up the wall and let herself hang down by her back legs from a peg, pretending to be dead. By and by, a mouse peeked his nose out, looking for food. The mouse paused when it saw the cat hanging there like a spider on a web. “Gosh!” the mouse cried. “You can’t fool us, you mangy cat! You can pretend to be a bag of bones hanging there until next Tuesday if you like, but don’t think you’ll catch us coming anywhere near you.” 
    
    If you are wise, you will not be fooled by the innocent actions of those you once found to be dangerous.
\end{quotation}

\newpage
\section{Wordlist}
\label{appendix:wordlist}

This was the 160-item wordlist used in this study. Participants saw the words very nearly as they are displayed here.

\begin{table}[h]
    \small
    \centering
    \resizebox*{!}{0.77\textheight}{%
        \begin{tabular}{c c c c} 
            push       & dawn         & Paul     & sparrow \\
            agony      & bow \& arrow & sugar    & yarn \\
            fulcrum    & wagon        & pawn     & row \\
            story      & pulpit       & peg      & bacon \\
            heritage   & control      & ignore   & ask \\
            pulley     & calm         & sew      & hairy \\
            corridor   & look         & board    & assault \\
            polar bear & smog         & Tom      & big \\
            carry      & whole        & flagrant & good \\
            thread     & low          & odd      & cook \\
            \\
            spot       & hook    & adult     & yuletide \\
            stood      & dog     & palm      & tulips \\
            vulture    & boss    & legacy    & through \\
            dodge      & dairy   & parrot    & rookie \\
            could      & throw   & go        & numeric \\
            bookie     & mow     & Coca-Cola & put \\
            dragon     & solve   & culprit   & pig \\
            regular    & show    & snag      & chocolate \\
            on         & holster & which     & new \\
            stroll     & bullet  & stew      & brag \\
            \\
            washing    & vague  & what   & sullen \\
            do         & cool   & would  & chew \\
            thrill     & moo    & star   & boot \\
            awe        & car    & clue   & dock \\
            quality    & swarm  & fall   & aunt \\
            two        & box    & school & argue \\
            wool       & book   & marble & golf \\
            dollar     & porch  & want   & throb \\
            goo        & salt   & dot    & holly \\
            floor      & plague & crook  & jaguar \\
            \\
            whipping   & sheriff  & loss       & reward \\
            bush       & black    & roof       & chorus \\
            majority   & exit     & pollen     & integrity \\
            boogie     & hooligan & course     & took \\
            corn       & toe      & inform     & paw \\
            chart      & Donald   & doe        & north \\
            spark      & deck     & narrate    & segment \\
            foot       & know     & volleyball & court \\
            student    & gullible & dolphin    & vary \\
            log        & doors    & rag        & Tuesday
        \end{tabular}
    }
    \label{tab:wordlist}
\end{table}





\newpage
\section{Minimal Pairs}
\label{appendix:minimal_pairs}

This was the list of 40 minimal pairs and 6 minimal triplets used in this study. Participants saw the words very nearly as they are displayed here.

\begin{table}[h]

    \begin{subtable}{\textwidth}
    \centering
        \begin{tabular}{ c c c c c}
            heal  & heel  &  \multicolumn{1}{p{2cm}}{\hspace{0in}} & here  & hear \\
            stairs & stares && mall & maul \\
            while  & wall   && hole & whole \\
            cot    & caught && Don  & dawn \\
            gull   & goal   && rule & roll \\
            \\
            meal     & mill    && hairy  & Harry \\
            mourning & morning && collar & caller \\
            jail     & gel     && bag    & beg \\
            hock     & hawk    && school & skull \\
            board    & bored   && sale   & sail \\
            \\
            hoarse  & horse  && odd    & awed \\
            daily   & deli   && dregs  & drags \\
            pair it & parrot && card   & cord \\
            hair    & hare   && leg    & lag \\
            vary    & very   && perish & parish \\
            \\
            Prego  & preggo && colt     & cult \\
            lord   & lard   && terrible & tear-able \\
            wail   & whale  && fairy    & ferry \\
            bowl   & bull   && bolder   & boulder \\
            holler & hauler && stole    & stool
        \end{tabular}
    \label{tab:minimal_pairs}
    \end{subtable}
    
    \bigskip
    
    \begin{subtable}{\textwidth}
        \centering
        \begin{tabular}{c c c}
                pole   & pull  & pool \\
                Mary   & merry & marry\\ 
                ore    & or    & are \\
                who’ll & hole  & hull \\
                full   & fool  & foal \\
                far    & four  & fore
        \end{tabular}
        \label{tab:my_label}
    \end{subtable}
    
\end{table}